Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-03-19 paper.qmd
Removed unnecessry code

trimesh_two_curvy_removed1_umap_with_data <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = tSNE1,
               y = tSNE2
             ),
             color = "#636363",
             alpha = 0.5,
             size = 0.5) +
  interior_annotation("b", sc_ltr_pos, cex = 1.5)
```

```{r}
#| label: best-fit-two-curvy-proj-umap1

@@

## Projection with Delauany triangulation

model_df1 <- dplyr::left_join(
  distance_two_curvy |> select(-distance), 
  projected_model_df, 
  by = c("from" = "ID"))

names(model_df1)[3:NCOL(model_df1)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_from")

model_df1 <- dplyr::left_join(model_df1, projected_model_df, by = c("to" = "ID"))
names(model_df1)[(2 + NCOL(projected_model_df)):NCOL(model_df1)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_to")

## With benchmark option 2

model_df2 <- dplyr::left_join(
  distance_df_small_edges_two_curvy3 |> select(-distance), 
  projected_model_df, 
  by = c("from" = "ID"))

names(model_df2)[3:NCOL(model_df2)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_from")

model_df2 <- dplyr::left_join(model_df2, projected_model_df, by = c("to" = "ID"))
names(model_df2)[(2 + NCOL(projected_model_df)):NCOL(model_df2)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_to")

@@

# axes_obj <- gen_axes(  
#   proj = projection * 2,
#   limits = 0.5,
#   axis_pos_x = -0.5,
#   axis_pos_y = -0.5,
#   axis_labels = names(scaled_two_curvy_data),
#   threshold = 0.03)
# 
# axes <- axes_obj$axes
# circle <- axes_obj$circle

two_curvy_proj_umap_first_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") + #31a354
  geom_point(
    size = 0.5,
    alpha = 0.2,
    color = "#636363") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.6, 0.6)) + 
  ylim(c(-0.6, 0.6)) +
  interior_annotation("c", sc_ltr_pos, cex = 1.5) +
  theme(
    legend.position = "none"
  )

two_curvy_proj_umap_first_model1_temp <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") + #31a354
  geom_point(
    size = 0.5,
    alpha = 0.2,
    color = "#636363") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.6, 0.6)) + 
  ylim(c(-0.6, 0.6)) +
  interior_annotation("c2", sc_ltr_pos) +
  theme(
    legend.position = "none"
  )

@@

two_curvy_proj_umap_all_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = true_model_df_proj, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#2b8cbe") +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") +
  geom_point(
    size = 0.5,
    alpha = 0.2,
    color = "#636363") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.6, 0.6)) + 
  ylim(c(-0.6, 0.6)) +
  interior_annotation("d", sc_ltr_pos, cex = 1.5) +
  theme(
    legend.position = "none"
  )

## Projection with Delauany triangulation

two_curvy_proj_umap_model_delaunay <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df1, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") + #31a354
  geom_point(
    size = 0.5,
    alpha = 0.2,
    color = "#636363") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.1, 0.1)) + 
  ylim(c(-0.1, 0.1)) +
  interior_annotation("a2", sc_ltr_pos) +
  theme(
    legend.position = "none"
  )

two_curvy_proj_umap_model_benchmark2 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df2, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") + #31a354
  geom_point(
    size = 0.5,
    alpha = 0.2,
    color = "#636363") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.1, 0.1)) + 
  ylim(c(-0.1, 0.1)) +
  interior_annotation("b2", sc_ltr_pos) +
  theme(
    legend.position = "none"
  )

@@

<!-- langevitour with pD model for two-curvy clusters -->
```{r}
#| label: best-fit-umap1
#| fig-cap: "We start with a tSNE layout (a) of non-linear clusters data ($n =  2000$ and $p = 4$), from which the \\gD{} wireframe is constructed (b) and then lifted to $4\\text{-}D$. A \\gD{} projection of the model fit from a tour through the $4\\text{-}D$ space is presented in c. Additionally, a \\gD{} projection of the fitted and true models (d) from the same tour demonstrates that the model accurately captures the non-linear structure of the data, although it does not fully represent the width of that structure. This highlights a key characteristic of tSNE, which compresses the data when applying tSNE. Furthermore, the sparse space observed in the middle of the tSNE layout is a result of (hyper-)parameter choice. The default is used. Video of the langevitour animation is available at <>."
#| fig-width: 10
#| fig-height: 10

nldr_two_curvy2 + 
  trimesh_two_curvy_removed1_umap_with_data +
  two_curvy_proj_umap_first_model1 +
  two_curvy_proj_umap_all_model1 +
  plot_layout(ncol = 2)
```

>>>>

trimesh_two_curvy_removed1_tsne_with_data <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = tSNE1,
               y = tSNE2
             ),
             color = "#636363",
             alpha = 0.5,
             size = 0.5) +
  interior_annotation("b", sc_ltr_pos, cex = 1.5)
```

```{r}
#| label: best-fit-two-curvy-proj-tsne

@@

@@

two_curvy_proj_tsne_first_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") + #31a354
  geom_point(
    size = 0.5,
    alpha = 0.2,
    color = "#636363") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.6, 0.6)) + 
  ylim(c(-0.6, 0.6)) +
  interior_annotation("c", sc_ltr_pos, cex = 1.5) +
  theme(
    legend.position = "none"
  )

@@

two_curvy_proj_tsne_all_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = true_model_df_proj, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#2b8cbe") +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") +
  geom_point(
    size = 0.5,
    alpha = 0.2,
    color = "#636363") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.6, 0.6)) + 
  ylim(c(-0.6, 0.6)) +
  interior_annotation("d", sc_ltr_pos, cex = 1.5) +
  theme(
    legend.position = "none"
  )

@@

<!-- langevitour with pD model for two-curvy clusters -->
```{r}
#| label: best-fit-tsne
#| fig-cap: "We start with a tSNE layout (a) of non-linear clusters data ($n =  2000$ and $p = 4$), from which the \\gD{} wireframe is constructed (b) and then lifted to $4\\text{-}D$. A \\gD{} projection of the model fit from a tour through the $4\\text{-}D$ space is presented in c. Additionally, a \\gD{} projection of the fitted and true models (d) from the same tour demonstrates that the model accurately captures the non-linear structure of the data, although it does not fully represent the width of that structure. This highlights a key characteristic of tSNE, which compresses the data when applying tSNE. Furthermore, the sparse space observed in the middle of the tSNE layout is a result of (hyper-)parameter choice. The default is used. Video of the langevitour animation is available at <>."
#| fig-width: 10
#| fig-height: 10

nldr_two_curvy2 + 
  trimesh_two_curvy_removed1_tsne_with_data +
  two_curvy_proj_tsne_first_model1 +
  two_curvy_proj_tsne_all_model1 +
  plot_layout(ncol = 2)
```
